---
title: "Top Songs of Spotify - 2023"
author: "Gaurav Surtani"
date: "2023-10-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Read data from CSV file:
```{r}
data <- read.csv("spotify-2023.csv")
```

### Print the first few rows of the data:
```{r}
head(data)
```

### Know the dimensions:
```{r}
dim(data)
```

### Rows and columns:
```{r}
nrow(data)
ncol(data)
```


### Checking for data quality issues:
```{r}
summary(data)
str(data)
head(data)
```

```{r}
missing_values <- sum(is.na(data))
cat("Number of missing values:", missing_values, "\n")

null_values <- sum(is.null(data))
cat("Number of null values:", null_values, "\n")
cat("Proportion of missing values:", mean(is.na(data)), "\n")
```

### Check the new dimensions of the cleaned data:
```{r}
fire_data <- na.omit(data)
dim(fire_data)
```

### Cleared the data with NA values
```{r}
missing_values_cleaned <- sum(is.na(fire_data))
missing_values_cleaned
```
 

#Introduction: 
I have chosen the Spotify dataset as it provides a comprehensive list of the most famous songs of 2023, along with a wealth of features about each song. I am curious to explore various aspects of popular music, including the impact of artists, the influence of song attributes on popularity, and how songs perform across different streaming platforms.  

## Proposal for Spotify Dataset Analysis

### Introduction

I have chosen the Spotify dataset as it provides a comprehensive list of the most famous songs of 2023, along with a wealth of features about each song. I am curious to explore various aspects of popular music, including the impact of artists, the influence of song attributes on popularity, and how songs perform across different streaming platforms.

### Data

- **Data Source**: The dataset is sourced from Kaggle[Link](https://www.kaggle.com/datasets/nelgiriyewithana/top-spotify-songs-2023/data) and Spotify and contains information about popular songs in 2023.

- **Data Collection**: The data was collected through a combination of sources, including Spotify's internal databases, music charts, and streaming statistics, through web scraping or API calls to gather additional information done in Kaggle

- **Cases**: Each row in the dataset represents a unique song. It provides detailed information about each song's attributes, popularity, and presence on various music platforms.

- **Variables**: (Referenced from Kaggle variable list)

   - `track_name`: Name of the song.
   - `artist(s)_name`: Name of the artist(s) of the song.
   - `artist_count`: Number of artists contributing to the song.
   - `released_year`: Year when the song was released.
   - `released_month`: Month when the song was released.
   - `released_day`: Day of the month when the song was released.
   - `in_spotify_playlists`: Number of Spotify playlists the song is included in.
   - `in_spotify_charts`: Presence and rank of the song on Spotify charts.
   - `streams`: Total number of streams on Spotify.
   - `in_apple_playlists`: Number of Apple Music playlists the song is included in.
   - `in_apple_charts`: Presence and rank of the song on Apple Music charts.
   - `in_deezer_playlists`: Number of Deezer playlists the song is included in.
   - `in_deezer_charts`: Presence and rank of the song on Deezer charts.
   - `in_shazam_charts`: Presence and rank of the song on Shazam charts.
   - `bpm`: Beats per minute, a measure of song tempo.
   - `key`: Key of the song.
   - `mode`: Mode of the song (major or minor).
   - `danceability_%`: Percentage indicating how suitable the song is for dancing.
   - `valence_%`: Positivity of the song's musical content.
   - `energy_%`: Perceived energy level of the song.
   - `acousticness_%`: Amount of acoustic sound in the song.
   - `instrumentalness_%`: Amount of instrumental content in the song.
   - `liveness_%`: Presence of live performance elements.
   - `speechiness_%`: Amount of spoken words in the song.

- **Type of Study**: This dataset is observational, as it provides information about songs and their attributes without any controlled experiments.

### Data Quality

I will perform the following checks:

- Check for missing data 
- Check any duplicates 


### Potential Use Cases ( Referenced : Kaggle Dataset )
1. **Music analysis**: Explore patterns in audio features to understand trends and preferences in popular songs.

2. **Platform comparison**: Compare the song's popularity across different music platforms.

3. **Artist impact**: Analyze how artist involvement and attributes relate to a song's success.

4. **Temporal trends**: Identify any shifts in music attributes and preferences over time.

5. **Cross-platform presence**: Investigate how songs perform across different streaming services.

Overall, this analysis aims to provide valuable insights into the world of popular music in 2023 and help stakeholders make informed decisions about their music-related endeavors.


##In Conclusion:
The Spotify dataset meets the condition as it includes:

1. **Categorical Variable**:
   - `track_name`: Name of the song.

2. **Continuous Numerical Variables**:
   - `danceability_%`: Percentage indicating how suitable the song is for dancing.
   - `valence_%`: Positivity of the song's musical content.
   - `energy_%`: Perceived energy level of the song.
   - `acousticness_%`: Amount of acoustic sound in the song.
   - `instrumentalness_%`: Amount of instrumental content in the song.